---
title: "Numerator relationship matrix"
output: html_document
date: "2025-02-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(package = "pedigreeTools")
```


## Numerator relationship matrix

The matrix that indicates the additive genetic relationship among individuals is called the numerator
relationship matrix (A). It is symmetric and its diagonal element for animal $i(a_{ii})$ is equal to $1 + F_i$, where $F_i$ is the inbreeding coefficient of animal $i$ (Wright, 1922). The diagonal element represent twice the probability that two gametes taken at random from animal $i$ will carry identical alleles by descent. The off-diagonal elements, $a_{ij}$, equals the numerator of the coefficient of relationship between animals $i$ and $j$. When multiplied by the additive genetic variance $xx$ is the covariance among breeding values. $A$ is twice the coefficient of parantages.

However, we can face memory bottlenecks, when we need to compute big pedigree...

```{r}
# Ax = b

# 1) Create a pedigree

ped = data.frame(iid = 1:7,
                 fid = c(0, 0, 1, 1, 3, 1, 5), 
                 mid = c(0, 0, 0, 2, 4, 4, 6))

(ped2 = pedigree(sire  = ped$fid, 
                dam   = ped$mid,
                label = ped$iid))
```

# Tabular method (recursive method) to compute A

1) If both parents $s$ and $d$ of animal $i$ are known:

- The diagonal elements $(a)_{ii}$ correspond to:
$a_{ii} = 1 + F_i = 1 + 0.5(a_{sd})$

- The off-diagonal element $a_{ji}$ correspond to:
$a_{ji} = a_{ij} = 0.5(a_{js} + a_{jd})$

2) If only one parent $s$ or $d$ of animal $i$ is known:

- The diagonal elements $(a)_{ii}$ correspond to:
$a_{ii} = 1$

- The off-diagonal element $a_{ji}$ correspond to:
$a_{ji} = a_{ij} = 0.5(a_{js/jd})$

3) If both parents are unknown and are assumed unrelated:

- The diagonal elements $(a)_{ii}$ correspond to:
$a_{ii} = 1$

- The off-diagonal element $a_{ji}$ correspond to:
$a_{ji} = a_{ij} = 0$

```{r}
A = getA(ped2)
A
```

# Thompson's method



TDT'

```{r}
(T = getT(ped2))
(D = getD(ped2))
```






