---
title: "Numerator relationship matrix"
output: html_document
date: "2025-02-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(package = "pedigreeTools") # R package to create and create different pedigree
                                   # features
```


# 1) Create a pedigree

```{r}
# Ax = b

# 1) Create a pedigree

ped = data.frame(iid = 1:7,
                 fid = c(0, 0, 1, 1, 3, 1, 5), 
                 mid = c(0, 0, 0, 2, 4, 4, 6))

(ped2 = pedigree(sire  = ped$fid, 
                dam   = ped$mid,
                label = ped$iid))
```

# 2) Recursive method to compute A

If both parents $s$ and $d$ of animal $i$ are known:

- The diagonal elements $(a)_{ii}$ correspond to:
$a_{ii} = 1 + F_i = 1 + 0.5(a_{sd})$

- The off-diagonal element $a_{ji}$ correspond to:
$a_{ji} = a_{ij} = 0.5(a_{js} + a_{jd})$

If only one parent $s$ or $d$ of animal $i$ is known:

- The diagonal elements $(a)_{ii}$ correspond to:
$a_{ii} = 1$

- The off-diagonal element $a_{ji}$ correspond to:
$a_{ji} = a_{ij} = 0.5(a_{js/jd})$

If both parents are unknown and are assumed unrelated:

- The diagonal elements $(a)_{ii}$ correspond to:
$a_{ii} = 1$

- The off-diagonal element $a_{ji}$ correspond to:
$a_{ji} = a_{ij} = 0$

```{r}
(A = getA(ped2))
```

# 3) Thompson's method

The relationship matrix can be expressed as $A = TDT^T$, where $T$ is a lower triangular matrix $D$ is a diagonal matrix. 

The matrix $T$ traces the flow of genes from one generation to the other; in other words, it accounts only for direct (parent-offspring) relationships. 

Rules for $i_{th}$: 

- $t_{ii} = 1$
- $t_{ij} = 0.5(t_{sj} + t_{dj})$ # If both parents (s and d) are known
- $t_{ij} = 0.5(t_{sj})$ # If only want parent (s) is known
- $t_{ij} = 0$ # If neither parent is known

```{r}
(T = getT(ped2))
```

The diagonal matrix $D$ is the variance and covariance matrix for Mendelian sampling. 


```{r}
(D = getD(ped2))
```







